Supplementary Report for the article *Functional diversity can facilitate the collapse of an undesirable ecosystem state*

### 2022-11-21 reports/supplement/supplement_0.9.1.Rmd

The rate of change in population densities is given by the growth rate - death rate [Bush et al 2017](https://www.nature.com/articles/s41467-017-00912-x). Growth rate is the growth rate in absence of any inhibition multiplied by the amount of inhibition.

@@

@@

For further information details of the functions used to model system, please see the original publication by Bush *et al.* (2018)

@@

Similarly, the complementarity effect and the sampling effect are important mechanisms by which diversity affects stability (Loreau & Mazancourt 2013; Loreau et al. 2021) and resilience (Steiner et al. 2006; Hughes & Stachowicz 2011). For example, diversity can stabilize aggregate ecosystem properties (e.g. community biomass) when species differ in their responses to environmental factors and therefore respond asynchronously to environmental fluctuations (temporal complementarity effect; Loreau & Mazancourt 2013). However, when considering ecosystem responses to directional environmental change or pulse disturbances rather than to environmental fluctuations, the sampling effect might be of particular importance (Steiner et al. 2006): more diverse systems might absorb a greater amount of change by having a higher chance of containing species with high environmental tolerance. In this scenario, the positive diversity effect on resilience is driven by one (or few) species. From a conservation perspective, however, one would nevertheless strive to maintain diversity and not only the most tolerant species because of the uncertainties associated with environmental change: we do not know with sufficient certainty which stressors will change in a given ecosystem and which traits will therefore be relevant in the future.

@@

## Do the observed diversity effects operate via the sampling effect?

We investigated if the diversity effects observed in our study were in line with the sampling effect, that is, driven entirely by the presence of the most tolerant strain. Specifically, we ran additional simulations where functional groups contained only the strain with greatest tolerance and lowest maximum growth rate. That is, for each of the 20 levels of trait variation, we made simulations where each functional group had only one strain, and its trait values corresponded with those of the most tolerant strain for the given level of trait variation. We also manipulated in which of the functional groups tolerance increased. For example, in a simulation to be compared with variation in only the SB group (and no variation in the CB or PB groups), the SB group contained the strain with highest tolerance, whereas CB and PB each contained the strain with mean tolerance. We then compared the tipping point positions resulting from this tolerance manipulation with the tipping point positions resulting from the diversity manipulation presented in the main text. In addition, we compared the dynamics of strains and substrates across the oxygen diffusivity gradient of the tolerance manipulation and the diversity manipulation. If the observed diversity effects are caused by the sampling effect, we expect there to be no difference between the tolerance manipulation and the diversity manipulation.

@@

For most of the explored combinations and amounts of functional diversity/tolerance the tipping point positions and strain dynamics did not differ between the diversity manipulation and the tolerance manipulation (Figure \@ref(fig:max-tol1)). However, in some cases there were effects, which are described in the follow sections.

@@

## Increasing tolerance vs. increasing diversity in PB

Whereas medium to high diversity in PB led to three states including one with coexistence of all three functional groups (Figure xy), presence of only the most tolerant PB strain led to only two stable states, namely the standard-pattern of either dominance of sulfur bacteria or of cyanobacteria (Figure xy). The most tolerant PB strain has a low intrinsic growth rate, leading to less consumption of sulfide and therefore to higher sulfide concentrations compared to a system that contains both slow- and fast-growing strains. Due to the higher sulfide concentrations in the system with only the most tolerant PB strain, cyanobacteria were not able to grow at low oxygen diffusivity levels.

@@

## Why only at high diversity/tolerance?

Why do these effects only occur at high diversity/tolerance? The likely reason is that at high diversities the most tolerant strain has particularly low growth rate, and so it may have the tolerance required to increase and persist, but not the growth rate required to dominate.

@@

@@

>>>>

The rate of change in population densities is given by the growth rate - death rate ([Bush et al 2017](https://www.nature.com/articles/s41467-017-00912-x)). Growth rate is the growth rate in absence of any inhibition multiplied by the amount of inhibition.

@@

Growth of cyanobacteria depended on one substrate (phosphorus), whereas growth of sulfate-reducing bacteria (SB) and phototrophic sulfur bacteria (PB) depended on two substrates (phosphorus and oxidised sulfur for SB, and phosphorus and sulfide for PB).

@@

For further information on the functions used to model the system, please see the original publication by [Bush et al 2017](https://www.nature.com/articles/s41467-017-00912-x).

@@

Similarly, the complementarity effect and the sampling effect are important mechanisms by which diversity affects stability (Loreau & Mazancourt 2013; Loreau et al. 2021) and resilience (Steiner et al. 2006; Hughes & Stachowicz 2011). For example, diversity can stabilize aggregate ecosystem properties (e.g. community biomass) when species differ in their responses to environmental factors and therefore respond asynchronously to environmental fluctuations (temporal complementarity effect; Loreau & Mazancourt 2013). However, the sampling effect might be of particular importance when considering ecosystem responses to directional environmental change or pulse disturbances rather than to environmental fluctuations (Steiner et al. 2006): more diverse systems might absorb a greater amount of change by having a higher chance of containing species with high environmental tolerance. In this scenario, the positive diversity effect on resilience is driven by one (or few) species. From a conservation perspective, however, one would nevertheless strive to maintain diversity and not only the most tolerant species because of the uncertainties associated with environmental change: we do not know with sufficient certainty which stressors will change in a given ecosystem and which traits will therefore be relevant in the future.

@@

## Investigating the mechanisms operating in our system

We investigated if the diversity effects observed in our study were in line with the sampling effect, that is, driven entirely by the presence of the most tolerant strain or if the less tolerant strains played a role as well. To this end, we ran additional simulations where functional groups contained only the strain with greatest tolerance and lowest maximum growth rate. That is, rather than increasing diversity we only increased tolerance. Specifically, for each of the 20 levels of trait variation, we made simulations where each functional group had only one strain, and its trait values corresponded with those of the most tolerant strain for the given level of trait variation. We also manipulated in which of the functional groups tolerance increased. For example, in a simulation to be compared with variation in only the SB group (and no variation in the CB or PB groups), the SB group contained the strain with highest tolerance, whereas CB and PB each contained the strain with mean tolerance. We then compared the tipping point positions resulting from this tolerance manipulation with the tipping point positions resulting from the diversity manipulation presented in the main text. In addition, we compared the strain dynamics across the oxygen diffusivity gradient of the tolerance manipulation and the diversity manipulation. If the observed diversity effects are caused by the sampling effect, we expect there to be no difference between the tolerance manipulation and the diversity manipulation.

@@

We found no difference between the diversity and the tolerance manipulation when trait values were low (Figure \@ref(fig:max-tol1)). When trait values were medium to high, however, we did find differences between the diversity and tolerance manipulation for most combinations except for diversity/tolerance in only CB and in SB-PB (Figure \@ref(fig:max-tol1)). We describe and discuss these differences in detail in the sections below. Briefly, in these scenarios less tolerant strains did play a role because of their high intrinsic growth rates. Specifically, absence of SB strains with high growth rate led to reduced production of sulfide and therefore to coexistence with cyanobacteria at low oxygen diffusivity. In contrast, absence of PB strains with high growth rate led to reduced consumption of sulfide and therefore precluded coexistence with cyanobacteria at low oxygen diffusivity (as observed when there was high variation in PB).

Taken together, at low amounts of trait variation the diversity effects were driven entirely by the most tolerant strain, in line with a sampling effect. At medium to high trait variation, however, both tolerant and less tolerant strains affected the results in most combinations. The reason was that in the two groups of sulfur bacteria the trade-off between tolerance and growth rate had ecosystem-level effects, i.e. effects on the sulfide concentration and therefore on the cyanobacteria.

@@

## Increasing tolerance vs. increasing diversity in PB and in PB-CB

Whereas medium to high diversity in only PB led to three states including one with coexistence of all three functional groups (Figure xy), presence of only the most tolerant PB strain led to only two stable states, namely the standard-pattern of either dominance of sulfur bacteria or of cyanobacteria (Figure xy). The most tolerant PB strain has a low intrinsic growth rate, leading to less consumption of sulfide and therefore to higher sulfide concentrations compared to a system that contains both slow- and fast-growing strains. Due to the higher sulfide concentrations in the system with only the most tolerant PB strain, cyanobacteria were not able to grow at low oxygen diffusivity levels.

For the same reasons we found three states on the trajectory of increasing oxygen diffusivity for medium to high simultaneous trait variation in PB and CB but not for the corresponding tolerance manipulation (data not shown).

@@

@@

Cardinale, B.J., Duffy, J.E., Gonzalez, A., Hooper, D.U., Perrings, C., Venail, P., et al. (2012). Biodiversity loss and its impact on humanity. Nature, 486, 59–67.

Cardinale, B.J., Matulich, K.L., Hooper, D.U., Byrnes, J.E., Duffy, E., Gamfeldt, L., et al. (2011). The functional role of producer diversity in ecosystems. Am. J. Bot., 98, 572–592.

Cardinale, B.J., Srivastava, D.S., Emmett Duffy, J., Wright, J.P., Downing, A.L., Sankaran, M., et al. (2006). Effects of biodiversity on the functioning of trophic groups and ecosystems. Nature, 443, 989–992.

Fargione, J., Tilman, D., Dybzinski, R., Lambers, J.H.R., Clark, C., Harpole, W.S., et al. (2007). From selection to complementarity: shifts in the causes of biodiversity–productivity relationships in a long-term biodiversity experiment. Proc. R. Soc. B Biol. Sci., 274, 871–876.

Hughes, A.R. & Stachowicz, J.J. (2011). Seagrass genotypic diversity increases disturbance response via complementarity and dominance. J. Ecol., 99, 445–453.

Loreau, M., Barbier, M., Filotas, E., Gravel, D., Isbell, F., Miller, S.J., et al. (2021). Biodiversity as insurance: from concept to measurement and application. Biol. Rev., 96, 2333–2354.

Loreau, M. & Hector, A. (2001). Partitioning selection and complementarity in biodiversity experiments. Nature, 412, 72–76.

Loreau, M. & Mazancourt, C. de. (2013). Biodiversity and ecosystem stability: a synthesis of underlying mechanisms. Ecol. Lett., 16, 106–115.

@@

Steiner, C.F., Long, Z.T., Krumins, J.A. & Morin, P.J. (2006). Population and Community Resilience in Multitrophic Communities. Ecology, 87, 996–1007.
